A forest is more than its trees: haplotypes and inferred ARGs

### 2024-11-22 paper/writeup.tex
edits

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the span over which ancestral haplotypes extend
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the XXX distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct,
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.

@@

We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
to take advantage of the extensive, stable, and well-documented set of open-source tools
available in \tskit{} \citep{tskit}.
So, our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived.
Some nodes are \emph{samples}, meaning that we know them from data. 
Typically, samples are exclusively leaves of $T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
over a segment of genome $[\ell_e, r_e)$.
(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

\subsection{Dissimilarity Function}
% * How to measure agreement that includes haplotypes
% * Definition and Algorithm
% * Supp fig: runtime ~ # trees, samples
% * fig: how add edges reduces discrepancy (RESULTS)
%       - compare to tsinfer
% * Show: how add edges reduces discrepancy (RESULTS)

If we begin with a tree sequence containing unary nodes,
it is straightforward to remove the portions of each node's span on which it is unary,
apply \algorithmref{alg:extend},
and quantify how much node span was correctly or incorrectly added.
However, we are also interested in whether \algorithmref{alg:extend}
improves \emph{inferred} tree sequences.
Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
that take into account haplotype identity,
we define a measure of \emph{dissimilarity} to quantify this.
The method is implemented in the \texttt{tsdate} package \citep{wohns2022unified}
as \comment{XXX}.

It is helpful to first describe what we compute in the simple case.
\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
for each node, the dark blue point is the total span in a ``true'' tree sequence
that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
in a way that included spans over which nodes are unary in the output
(by setting \texttt{coalescing\_segments\_only} to False).
The green point is the span of the node after \emph{simplifying} the tree sequence
\citep{kelleher2018efficient},
which removes all unary nodes from all trees.
The red point
is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
\figref[B]{fig:node-spans} shows how much of this span is incorrect:
for each node, it gives the amount of span
that is incorrect (i.e., not present in the original tree sequence)
over the total amount of span added after running \algorithmref{alg:extend}.
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

@@

Concretely, our method can be applied to any two tree sequences,
 as in the case of \figref{fig:conceptual_discrepancy},
 where we consider the pair of trees which exist at that break point.
However, suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings of nodes in $\T_1$ to nodes in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

@@

The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
but not times.
If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
to also compare times,
for instance by returning the weighted root-mean-squared error on inferred times as
\begin{align}\label{eqn:wrmse}
    \text{wRMSE}_t(\T_1, \T_2)
    = \sqrt{\frac{
        \sum_{n \in N_1} s(\T_1,n) \left(\log\left(1+ t^{(1)}_n\right) - \log\left(1+ t^{(2)}_{\alpha(n)}\right) \right)^2 
    }{
        \sum_{n \in N_1} s(\T_1,n)
    } } .
\end{align}
The mean is computed weighting by node span, so that a dating error
is more serious for a node with a longer span.

The implementation of this method, called \emph{tree\_dissimilarity} \comment{change?}, in \texttt{tsdate}
produces three values after computation: 
the relative dissimilarity, weighted root-mean-square-error, and true proportion represented. 
The \emph{relative dissimilarity} is defined to be the dissimilarity proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\overline{\dis}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}
 	=1-\frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}.
\end{align}
The \emph{weighted root-mean-square-error} is given in \eqnref{eqn:wrmse}.
The \emph{true proportion represented} is similarity between two trees relative 
to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\overline{\similarity}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_2}s(\T_2,n)}.
\end{align}
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
We compute the tree\_dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.

%%%%%%%%%%%%
%\comment{PREVIOUS VERSION: retrieve bits from below?}
%
%Inspired by Robinson-Foulds we define a function, called a \textit{discrepancy function} (see \algorithmref{alg:disc}), on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
%Let $N_{T_1}$ and $N_{T_2}$ be the set of nodes on $\T_1$ respectively $\T_2$.
%For a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
%Each node $n\in\T$ has a given position along the genome and time, 
%so we can write each node as a tuple $n = (t_n, L_n)$ with position $L_n\subset \left[0,L\right)$ and time $t_n\in \R^{\geq 0}$.
%Computing the discrepancy between a random pair of nodes is not helpful, 
%so we must first identify which pairs of nodes would be most identical to one another or a `best match'.
%These `best matching' nodes should be close in terms of time and have equal descendant sample sets,
%ie. $s(i) = s(j)$ where $s$ is the function to compute the descendant sample set of $n_1$.
%
%To find a best match for each node $i\in N_{T_1}$ \comment{(Should I do this by mentioning clades?)}
%we compute the \textit{shared span} between node $i\in N_{T_1}$ and every node in $N_{T_2}$.
%Then, the \textit{shared span} between two tree sequences, denoted $ss(\T_1,\T_2)$, is a $|N_{T_1}|\times |N_{T_2}|$ matrix $A$
%where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$.
%If a pair $(i,j)$ has equal descendant sample sets, $s(i) = s(j)$
%then $a_{ij} = |L_i \cap L_j|$.
%If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
%Matrix $A$ determines the \textit{best match} for every node $i\in N_{T_1}$
%via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_{T_2}} a_{ij}$.
%If this maximum is not unique, we use time difference to determine the best match pair.
%We then define the \textit{discrepancy} between $\T_1$ and $\T_2$ as 
%one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
%
%We note that the discrepancy function is a metric for detecting non-similarity between two tree sequences. However, we note that this function is not a metric in the mathematical sense, as it is not symmetric, nor is it reflexive. For example, given a tree sequence $\T$ and its simplification $\mathbb{S}$ and $\T\neq\mathbb{S}$ we have $\operatorname{discrepancy}(\mathbb{S}, \T)=0$. Additionally, given \figref{fig:conceptual_discrepancy}, $\operatorname{discrepancy}(T1, T2) \neq \operatorname{discrepancy}(T2, T1)$ \comment{actually compute this. We may just do it in this section instead of the appendix?}
% 
%\comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets.
        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. On the rest of the genome, all of nodes match with their identical counterpart.
        This makes the dissimilarity $\dis(T1,T2)=3$ and the \emph{relative similarity} $\overline{\dis}(T1,T2)=\frac{3}{46}$.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

%\comment{Delete Discrepancy Algorithm!!}
% 
% \begin{algorithm2e}[!ht]
% \SetStartEndCondition{ }{}{}%
% \SetKwProg{Fn}{def}{\string:}{}
% \SetKwFunction{Range}{range}%%
% \SetKw{KwTo}{in}
% \SetKwFor{For}{for}{\string:}{}%
% \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
% \SetKwFor{While}{while}{:}{fintq}%
% \SetKw{Break}{break}
% \newcommand\forcond{$i$ \KwTo\Range{$n$}}
% \SetKwComment{Comment}{}
% \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
% \DontPrintSemicolon\LinesNotNumbered
% 
% \caption{Discrepancy Function}\label{alg:disc}
% \input{$(T_1,T_2)$}
% \output{tree discrepancy, root-mean-square error}
% 
% \Fn{Shared Spans($T_1,T_2$)}{
% \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
% 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
% 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
% 		Define $a_{ij}=|L_i\cap L_j|$.
% 	}
% 	\Else{Define $a_{ij}=0$.}
% Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
% }
% \KwRet{$A$}
% }
%\BlankLine
%\Fn{Discrepancy ($T_1,T_2$)}{
%\For{each node $i\in N_{T_1}$}{
%	Find nodes in $N_{T_2}$ which have the highest matching span.\;
%	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
%	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
%	Define $\alpha_i = a_{ij}$.\;
%	\BlankLine
%}
%$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_1)\right)}$$\;
%\BlankLine\BlankLine
%$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_1)\right)}}$$\;
%\BlankLine
%$$true\ proportion = \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname(tr)\left( Shared Spans(T_2, T_2)\right)}$$\;
%\BlankLine\BlankLine
%\KwRet{tree discrepancy, rmse, true proportion}
%}
%\end{algorithm2e}

\subsection{Simulations}

@@

We used simulations of several scenarios.
To include the effects of heterogeneous recombination rate,
in some we used stdpopsim \citep{stdpopsim} 
to simulate chromosome 1 of \textit{Canis familiaris}
using the CanFam3 genetic map from \citet{campell2016}.
\emph{``Constant dog''} simulations simulated this chromosome in a 
population of (constant) size $10^4$.
\emph{``Expanding dog''} simulations were similar, but
used a discrete-time Wright--Fisher model to simulate
a small population of 100 that expanded to 1,000 individuals ten generations ago,
which then doubled every generation to reach 512,000 individuals.
All jobs used for recording runtime were run on an Intel Xeon Gold 6148 processor.
To observe the accuracy of our method, we compute node spans for a correctly and 
incorrectly added for a tree sequence simulated with
 1,000 samples over a genome of length $10^8$.
 Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
 using our dissimilarity method. 
 These tree sequences range from 10 to 1,000 samples with genome length $5\times 10^7$
 as well as having 1,000 samples and ranging the genome length from $10^6$ to $5\times 10^7$.
All jobs used for recording node span and (dis)similarity for true and inferred ARGs
 were run on an Intel Core i5-1035G4 processor.

@@

This reduction in edges can also lead to a reduction in computation time.
Indeed, \figref{fig:speed_and_edges} shows that computation time is reduced by 10--20\%
for a typical statistic (here, Tajima's $D$),
computed in an efficient iterative manner along the genome as implemented in \tskit.
This is actually somewhat surprising:
as described in \citet{ralph2020efficiently},
each edge added or removed requires an update
to the state of the parent node \emph{and every node above it in the tree}.
So, even if an edge extends for a long segment of genome,
something has to happen to it every time the subtree below it changes,
which is not affected by extending haplotypes.

@@

\begin{figure}[bht]
	\includegraphics{newplots_wo_ee/figure-4.pdf}
	\caption{
        The effect of extending haplotypes on per-node spans
        in an ARG simulated with $10^4$ \comment{diploid?} samples on a sequence of length $10^8$.
        \textbf{(A)}
        Distribution of total amount of span added across nodes by \algorithmref{alg:extend};
        note the log scale on the $y$ axis.
        \textbf{(B)}
        Amount of incorrectly added span, plotted against total span, by node.
        95\% of nodes have no incorrect span; of the remainder,
        nearly all have less than 5\% incorrectly added;
        see Supplementary \figref{fig:incorrect_ratio}.
        Plots \textbf{(C)} and \textbf{(D)} show total spans per node,
        ordered by total span in the original ARG (which includes unary nodes).
        Dark blue dots in both show the spans in this original ARG.
        In \textbf{(C)}, lighter green dots show
        span after removing unary spans using \textit{simplify},
        while in \textbf{(D)},
        lighter red dots show span after extending haplotypes,
        i.e., applying \algorithmref{alg:extend} to the simplified ARG.
    }
	\label{fig:node-spans}
\end{figure}

These statistics are also reflected at the genomic scale, using measures of dissimilarity.
\figref{fig:dissimilarity} shows in purple total amounts of span removed by simplification
and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
(Lines in green involve re-inference of the ARG; discussed next.)
The top row shows the proportion of the given ARG that does not match the original,
showing that the total amount of mis-matching span
produced by extending haplotypes is very small ($\approx 1\%$).
(Simplification does not produce dissimilarity.)
The bottom row shows the proprtional of the original ARG that is represented in the given ARG.
This shows that a large proportion of the ARG can be removed by simplifiction,
indicating that coalescent nodes are unary
over a substantial portion of their spans.
(Since the \texttt{simplify} operations removes these unary portions of haplotypes,
the ``simplified'' line (``S'') on the right of Figure~\ref{fig:dissimilarity} shows
the proportion of nodes' spans on which they are not unary.)
However, \algorithmref{alg:extend} can correctly replace most of these portions of haplotypes,
especially with larger sample sizes and sequence lengths
(the ``simplified--extended'' line; ``SE'').
For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
haplotypes are unary on about half their spans (on average),
and extending haplotypes can infer more than half of this span from coalescent information only.

>>>>

In this paper, we study various aspects of haplotype identity in ARGs.
First, we describe a deterministic algorithm that
extends the span over which ancestral haplotypes extend
using the principle that intermediate nodes in inheritance paths
should remain unchanged when possible.
These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
To quantify how accurate the new information is,
we define and describe how to compute new measures of (dis-)agreement between ARGs
that are motivated by the Robinson-Foulds distance between trees
but account for haplotype identity.
These measures show that the vast majority of these extended haplotypes are correct,
and that substantial information about haplotypes is contained in these nodes
in inferred trees as well.

@@

We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
to take advantage of the extensive, stable, and well-documented set of open-source tools
available in \tskit{} \citep{tskit}.
So, our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived.
Some nodes are \emph{samples}, meaning that we know them from data. 
Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
over a segment of genome $[\ell_e, r_e)$.
(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

\subsection{Dissimilarity between ARGs}
% * How to measure agreement that includes haplotypes
% * Definition and Algorithm
% * Supp fig: runtime ~ # trees, samples
% * fig: how add edges reduces discrepancy (RESULTS)
%       - compare to tsinfer
% * Show: how add edges reduces discrepancy (RESULTS)

If we begin with a tree sequence containing unary nodes,
it is straightforward to remove the portions of each node's span on which it is unary,
apply \algorithmref{alg:extend},
and quantify how much node span was correctly or incorrectly added.
However, we are also interested in whether \algorithmref{alg:extend}
improves \emph{inferred} tree sequences.
Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
that take into account haplotype identity,
we define a measure of \emph{dissimilarity} to quantify this.
The method is implemented in the \texttt{tscompare} package.

It is helpful to first describe what we compute in the simple case.
\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
for each node the total span in a ``true'' tree sequence, 
given by the line,
that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
in a way that included spans over which nodes are unary in the output
(by setting \texttt{coalescing\_segments\_only} to False).
\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
\citep{kelleher2018efficient},
which removes all unary nodes from all trees
thus reducing the node spans for some nodes in the sequence.
In \figref[D]{fig:node-spans}
is the span of the node after applying \algorithmref{alg:extend}
to the simplified tree sequence.
\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
\figref[B]{fig:node-spans} shows how much of this span is incorrect:
for each node, it gives the amount of span
that is incorrect (i.e., not present in the original tree sequence)
over the total amount of span added after running \algorithmref{alg:extend}.
The values in these plots (perhaps summed over nodes)
therefore quantify how much the ancestral haplotypes represented by each node
were correctly, and incorrectly, extended.
They also tell us how much of the relationships in the original tree sequence
are represented in the result.

@@

Concretely, our method can be applied to any two tree sequences,
 as in the case of \figref{fig:conceptual_discrepancy},
 where we consider the pair of trees which exist at that break point.
However, suppose that the two tree sequences have the same set of breakpoints
between trees,
so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
For a node $n$ and tree $T$
let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
define
\begin{align*}
    m(n_1, n_2)
    =
    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
\end{align*}
which is the total span over which the samples below $n_1$ in $\T_1$
matches the samples below $n_2$ in $\T_2$.
Also let
\begin{align*}
    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
\end{align*}
the total span that node $n$ is present in the marginal trees.
The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
\end{align*}
where the maximum is over all mappings of nodes in $\T_1$ to nodes in $\T_2$.
(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
it's \emph{best matching node} in $\T_2$ as
\begin{align*}
    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
\end{align*}
so that
\begin{align*}
    \similarity(\T_1, \T_2)
    =
    \sum_{n \in N_1} m(n, \alpha(n)) .
\end{align*}
If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
(and if \emph{this} is not unique, we pick an arbitrary one) --
however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
\begin{align*}
    \dis(\T_1, \T_2)
    =
    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
    = 
    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
\end{align*}
which is the proportion of total span of all nodes in $\T_1$
for which their descendant samples do \emph{not} match those of their best match in $\T_2$.

@@

The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
but not times.
If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
to also compare times,
for instance by returning the weighted root-mean-squared error on inferred times as
\begin{align}\label{eqn:wrmse}
    \text{wRMSE}_t(\T_1, \T_2)
    = \sqrt{\frac{
        \sum_{n \in N_1} s(\T_1,n) \left(\log\left(1+ t^{(1)}_n\right) - \log\left(1+ t^{(2)}_{\alpha(n)}\right) \right)^2 
    }{
        \Vert{\T_1}\Vert
    } } .
\end{align}
The mean is computed weighting by node span, so that a dating error
is more serious for a node with a longer span.

The implementation of this method in \texttt{tscompare}
additionally produces relative values:
the ancestral recombination graph robinson-foulds value (relative dissimilarity)  and true proportion represented. 
The \emph{ancestral recombination graph robinson-foulds} (ARF) is defined to be the dissimilarity proportional
 to the total span of nodes in $\T_1$,
 \begin{align}\label{eqn:dissimilarity}
 	\operatorname{ARF}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\Vert \T_1 \Vert}
 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert}.
\end{align}
The \emph{true proportion represented} (TPR) is similarity between two trees relative 
to the total span of nodes in $\T_2$,
\begin{align}\label{eqn:true-proportion}
	\operatorname{TPR}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}
	{\Vert \T_2\Vert}.
\end{align}
The outputs framed as proportions relative to one of the given tree sequences is more easily understood
for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
We compute dissimilarity and ARF of two tree sequences in \figref{fig:conceptual_discrepancy}.

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets.
        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. Thus the maximal mapping for 4 should be to itself. On the rest of the genome, all of nodes match with their identical counterpart.
        This makes the dissimilarity $\dis(T1,T2)=3$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Simulations}

@@

We used simulations of several scenarios.
To include the effects of heterogeneous recombination rate,
in some we used stdpopsim \citep{stdpopsim} 
to simulate chromosome 1 of \textit{Canis familiaris}
using the CanFam3 genetic map from \citet{campell2016}.
\emph{``Constant dog''} simulations simulated this chromosome in a 
population of (constant) size $10^4$.
\emph{``Expanding dog''} simulations were similar, but
used a discrete-time Wright--Fisher model to simulate
a small population of 100 that expanded to 1,000 individuals ten generations ago,
which then doubled every generation to reach 512,000 individuals.
All jobs used for recording runtime were run on an Intel Xeon Gold 6148 processor.
 Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
 using our dissimilarity methods. 
 These tree sequences have a population of 1,000 and recombination rate of $10^{-8}$ with samples ranging from 10 to 1,000 and genome length ranging from $10^6$ to $5\times 10^7$.

@@

This reduction in edges can also lead to a reduction in computation time.
Indeed, \figref{fig:speed_and_edges} shows that computation time is reduced by 10--20\%
for a typical statistic (here, Tajima's $D$),
computed in an efficient iterative manner along the genome as implemented in \tskit.
This may at first be somewhat surprising:
This makes sense because
as described in \citet{ralph2020efficiently},
each edge added or removed requires an update
to the state of the parent node and nodes above it,
and extending haplotypes yields a tree sequence with fewer edge removals and insertions.

@@

\begin{figure}[!bht]
	\includegraphics{newplots_wo_ee/figure-4.pdf}
	\caption{
        The effect of extending haplotypes on per-node spans
        in an ARG simulated with $10^4$ diploid samples, $10^4$ population, and 
        recombination rate of $10^{-8}$ on a sequence of length $10^8$.
        \textbf{(A)}
        Distribution of total amount of span added across nodes by \algorithmref{alg:extend};
        note the log scale on the $y$ axis.
        \textbf{(B)}
        Amount of incorrectly added span, plotted against total span, by node.
        95\% of nodes have no incorrect span; of the remainder,
        nearly all have less than 5\% incorrectly added;
        see Supplementary \figref{fig:incorrect_ratio}.
        Plots \textbf{(C)} and \textbf{(D)} show total spans per node,
        ordered by total span in the original ARG (which includes unary nodes).
        Dark blue dots in both show the spans in this original ARG.
        In \textbf{(C)}, lighter green dots show
        span after removing unary spans using \textit{simplify},
        while in \textbf{(D)},
        lighter red dots show span after extending haplotypes,
        i.e., applying \algorithmref{alg:extend} to the simplified ARG.
    }
	\label{fig:node-spans}
\end{figure}

These statistics are also reflected at the genomic scale, using measures of dissimilarity.
Line ``SE'' in \figref{fig:dissimilarity} shows total amounts of span removed by simplification
and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
(Lines labeled with ``I'' involve re-inference of the ARG; discussed next.)
The top row shows the proportion of the given ARG that does not match the original (ARF),
showing that the total amount of mis-matching span
produced by extending haplotypes is very small ($\approx 1\%$).
(Simplification does not produce dissimilarity.)
The bottom row shows the proprtional of the original ARG that is represented in the given ARG (TRP).
This shows that a large proportion of the ARG can be removed by simplifiction,
indicating that coalescent nodes are unary
over a substantial portion of their spans.
(Since the \texttt{simplify} operations removes these unary portions of haplotypes,
the ``simplified'' line (``S'') on the right of Figure~\ref{fig:dissimilarity} shows
the proportion of nodes' spans on which they are not unary.)
However, \algorithmref{alg:extend} can correctly replace most of these portions of haplotypes,
especially with larger sample sizes and sequence lengths
(the ``simplified--extended'' line; ``SE'').
For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
haplotypes are unary on about half their spans (on average),
and extending haplotypes can infer more than half of this span from coalescent information only.

\begin{figure}[!hbt]
	\begin{center}
        \includegraphics[width=\textwidth]{figure5.pdf}
	\end{center}
    \caption{
        Accuracy and sensitivity of extended haplotypes
        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
        and sequence lengths, with 1,000 diploid samples (bottom).
        For each, a simulated ARG containing unary haplotype spans
        was (i) \emph{simplified}, removing the unary spans,
        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
        then each of these had its haplotypes \emph{extended}.
        The inferred, then simplified, ARG is also shown.
        \textbf{Top row:} ARF, the dissimilarity to the true ARG,
        as proportion of haplotypes that are not represented in the true ARG
        (\eqnref{eqn:dissimilarity}).
        \textbf{Bottom row:} TPR, agreement to the true ARG
        as proportion of the true ARG that is represented.
        }     
    \label{fig:dissimilarity}
\end{figure}
